### 2021-10-28 papers/modeling-of-combined-lfr-csp-sco2-paper/content.tex
backing up

\replaced[comment={Abbreviation defined [4,1]}]{Supercritical carbon dioxide (sCO$_2$).}{Supercritical CO$_2$} Brayton cycles are promising cycle configurations offering higher efficiencies, compact design, and reduced turbomachinery cost while operating with non toxic working fluid. Various sCO$_2$ Brayton cycles have been modeled with the recompression cycle having efficiency advantages over other proposed cycle arrangements \cite{turchi_2013,ahn_2014,wang_2018}. The literature shows that the recompression cycle can reach efficiencies of 50\% in some scenarios (with turbine inlet temperatures in the 650-700 $^{\circ}$C range) \cite{turchi_2013,wright_2009}, allows these cycles to be a competitive alternative to steam Rankine and air Brayton cycles over a range of temperatures. Due to the benefits of sCO$_2$ Brayton cycles, the United States Department of Energy is investigating these conversion cycles for use with heat sources including nuclear and solar \cite{doe_2012}. Multiple project funding opportunities are established with the National Energy Technology Laboratory offering 144 million dollar award for demonstration and performance verification of a sCO$_2$ Brayton cycle \cite{netl_2016} and the Office of Energy Efficiency \& Renewable Energy offering 2.6 million dollar reward in their Brayton Energy project for a integrated \replaced[comment={Abbreviations defined [4,1]}]{concentrating solar power (CSP) receiver, thermal energy storage (TES),.}{CSP receiver, TES,} and power block in one sCO$_2$ Brayton system \cite{seto_2015}. Utilizing complementary technologies, specifically solar concentrating power and lead-cooled fast reactors, can offset the drawbacks of each. \replaced[comment={Drawbacks of CSP and LFR technology elaborated on [2,2]}]{These drawbacks include CSP dependency on weather conditions and time of day, while the LFR has a large system inertia on start-up and ramping while reducing strain on system components when consistently operating.}{} Coupling these technologies into an interconnected cycle allows for consistent generation, independent of weather or time of day, and thermal storage for high dispatchability during high grid demand periods. 
A CSP has an array of mirrors concentrating solar rays towards a \replaced[comment={Furthering explanation of the CSP [3,1]}]{tower receiver, transferring the energy into a solar salt, and storing the solar salt in tanks. CSP systems require direct sunlight to operate at full power therefore causing variability based on weather conditions and time of day.}{receiver to generate thermal energy therefore causing a dependency on weather conditions and time of day.} Previous research has modeled CSP technology as the thermal source for Brayton sCO$_2$ cycles with promising results in efficiency gains and high temperature thermal energy storage \cite{turchi_2013, iverson_2013, ho_2015, wang_2018, neises_2020}. Thermal energy gathered from the CSP is stored in thermal energy storage which is dispatched into the sCO$_2$ Brayton cycle when grid demand increases. With a comparable temperature to the hot TES, 560 $^{\circ}$C, a \replaced[comment={Abbreviation defined [4,1]}]{lead-cooled fast reactor (LFR).}{LFR} is capable of transferring heat to thermal storage for later dispatch. An LFR uses fission reactions to heat a heat transfer fluid, in this case liquid lead, to temperatures of 650 $^{\circ}$C. Nuclear sCO$_2$ Brayton cycles have been studied with similar gains in efficiency seen in CSP sCO$_2$ Brayton cycles \cite{ dostal_2004, luo_2020}. Physical testing of simple sCO$_2$ Brayton cycles has been performed at Sandia National Laboratories \cite{wright_2011} and in Korea \cite{cha_2016}, with at least two commercial companies (Echogen, Net Power) having progressed to building pilot cycles  \cite{held_2015,fetvedt_2016}.

@@

Counter-flow heat exchangers are modeled with two fluids flowing in opposite directions exchanging heat from the hot side to the cold side. \replaced[comment={Improving counter-flow hx explanation [1,1]}]{The smallest temperature difference of the hot and cold flows on either the low or high end of the heat exchanger is defined as the approach temperature of the counter-flow heat exchanger.}{The temperatures of the hot and cold flows on either side of the heat exchanger have a temperature difference known as an approach temperature.} A diagram showing a simplified counter-flow heat exchanger is illustrated in Figure \ref{counter-flow-hx}.

@@

Where $C_{avg}$ is the average capacitance rate across the hot or cold stream and $c_{in}$ and $c_{out}$ is the specific heat at the inlet and outlet respectively. \replaced[comment={Improving counter-flow hx eplanation [1,1]}]{Specific heat is found using library correlations, with the average capacitance rate assumed to be constant during the analysis. Assuming constant average capacitance rate is suitable for most engineering purposes, especially when there is uncertainty associated with other parameters \cite{nellis_klein_2008}.}{Specific heat is found using library correlations.} Once both average capacitances are calculated for the hot and cold streams, one has a larger value, $\dot{C}_{max}$, and one has a smaller value,  $\dot{C}_{min}$. These maximum and minimum values are used to find the capacitance ratio, $CR$, in Equation \ref{cap_ratio}.

@@

@@

Where the left hand side, $\dot{m}$, $h_{inlet}$, and $\dot{Q}_{LFRHX}$, is the energy into the flow and the right hand side, $\dot{m}$ and $h_{outlet}$, is the energy brought out from the flow of sCO$_2$. The amount of energy transferred into the cycle, $\dot{Q}_{LFRHX}$, is set at $950$ MW, and outlet temperature of the sCO$_2$ from the LFR HX is set at a value of $595^{\circ}$C. The outlet temperature of the LFR is specified because of high temperature material limits on the LFR lead side. The low temperature side is allowed to vary over a range of values with some considerations. The lead flow velocity is limited by the erosion of the fuel, the slower the lead flow velocity reduces fuel erosion and therefore leads to a more desirable compact design. \replaced[comment={Edited sentence for clarity [2,3]}]{Restricting the lead flow velocity, and therefore lead mass flow rate, while lowering the inlet sCO$_2$ temperature increases the temperature difference across the LFR HX allowing for a higher, more desirable, LFR power output.}{Restricting the lead flow velocity, and therefore lead mass flow rate, leads to a higher LFR power output when the inlet sCO$_2$ temperature is reduced.} LFR sCO$_2$ inlet temperature has a lower bound of $340^{\circ}$C before the lead begins to freeze, which is operationally unacceptable. When the inlet temperature of sCO$_2$ is increased the temperature difference across the LFR is decreased leading to an increase in power conversion cycle thermodynamic efficiency but a reduction in LFR power below $950$ MW. There is a compromise between high LFR efficiency and LFR power, therefore a temperature of $400^{\circ}$C for the sCO$_2$ inlet temperature is the optimal value provided by Westinghouse.

@@

When grid demand for electrical power increases, a series of operating modes are activated. During the highest demand times, cycle operation focuses on maximum electrical generation. This is achieved through the C2S being turned off for direct electrical production from the LFR and the hot TES is discharging heat through the CSP HX for electrical production in the sCO$_2$ Brayton cycle. \replaced[comment={Edited sentence, CSP still charges the hot TES during low grid demand periods [2,4]}]{As grid demand diminishes, electrical generation is ramped down by decreasing heat input through the CSP HX. This continues until no heat is being dispatched from the solar salt to the sCO$_2$, and all produced thermal energy from the CSP is put into charging the hot TES.}{As grid demand diminishes, CSP HX ramps down heat extraction until no power is being dispatched through the salt and the hot TES begins charging.} During this process, the LFR gradually adds a larger fraction of heat input to the TES through C2S. This process continues until no electrical production is occurring in the cycle and all heat is stored in TES for later use.

>>>>

\replaced[comment={Abbreviation defined [4,1]}]{Supercritical carbon dioxide (sCO$_2$).}{Supercritical CO$_2$} Brayton cycles are promising cycle configurations offering higher efficiencies, compact design, and reduced turbomachinery cost while operating with non toxic working fluid. Various sCO$_2$ Brayton cycles have been modeled with the recompression cycle having efficiency advantages over other proposed cycle arrangements \cite{turchi_2013,ahn_2014,wang_2018}. The literature shows that the recompression cycle can reach efficiencies of 50\% in some scenarios (with turbine inlet temperatures in the 650-700 $^{\circ}$C range) \cite{turchi_2013,wright_2009}, allows these cycles to be a competitive alternative to steam Rankine and air Brayton cycles over a range of temperatures. Due to the benefits of sCO$_2$ Brayton cycles, the United States Department of Energy is investigating these conversion cycles for use with heat sources including nuclear and solar \cite{doe_2012}. Multiple project funding opportunities are established with the National Energy Technology Laboratory offering 144 million dollar award for demonstration and performance verification of a sCO$_2$ Brayton cycle \cite{netl_2016} and the Office of Energy Efficiency \& Renewable Energy offering 2.6 million dollar reward in their Brayton Energy project for a integrated \replaced[comment={Abbreviations defined [4,1]}]{concentrating solar power (CSP) receiver, thermal energy storage (TES),.}{CSP receiver, TES,} and power block in one sCO$_2$ Brayton system \cite{seto_2015}. Utilizing complementary technologies, specifically solar concentrating power and lead-cooled fast reactors, can offset the drawbacks of each. \replaced[comment={Drawbacks of CSP and LFR technology elaborated on [2,2]}]{These drawbacks include CSP dependency on weather conditions and time of day, while the LFR in isolation does no incoporate thermal energy storage for meeting peak demand}{} Coupling these technologies into an interconnected cycle allows for consistent generation, independent of weather or time of day, and thermal storage for high dispatchability during high grid demand periods. 
A CSP has an array of mirrors concentrating solar rays towards a \replaced[comment={Furthering explanation of the CSP [3,1]}]{tower receiver, transferring the energy into a solar salt, and storing the solar salt in tanks. CSP systems require direct sunlight to operate at full power therefore causing variability based on weather conditions and time of day.}{receiver to generate thermal energy therefore causing a dependency on weather conditions and time of day.} Previous research has modeled CSP technology as the thermal source for Brayton sCO$_2$ cycles with promising results in efficiency gains and high temperature thermal energy storage \cite{turchi_2013, iverson_2013, ho_2015, wang_2018, neises_2020}. Thermal energy gathered from the CSP is stored in thermal energy storage which is dispatched into the sCO$_2$ Brayton cycle when grid demand increases. With a comparable temperature to the hot TES, 560 $^{\circ}$C, a \replaced[comment={Abbreviation defined [4,1]}]{lead-cooled fast reactor (LFR).}{LFR} is capable of transferring heat to thermal storage for later dispatch. An LFR uses fission reactions to heat a heat transfer fluid, in this case liquid lead, to temperatures of 650 $^{\circ}$C. Nuclear sCO$_2$ Brayton cycles have been studied with similar gains in efficiency seen in CSP sCO$_2$ Brayton cycles \cite{ dostal_2004, luo_2020}. Physical testing of simple sCO$_2$ Brayton cycles has been performed at Sandia National Laboratories \cite{wright_2011} and in Korea \cite{cha_2016}, with at least two commercial companies (Echogen, Net Power) having progressed to building pilot cycles  \cite{held_2015,fetvedt_2016}.

@@

Counter-flow heat exchangers are modeled with two fluids flowing in opposite directions exchanging heat from the hot side to the cold side. \replaced[comment={Improving counter-flow hx explanation [1,1]}]{The smallest temperature difference of the hot and cold flows on either the low or high end of the heat exchanger is defined as the approach temperature of the counter-flow heat exchanger and a calculation is performed to identify whether the hot end of cold end is limiting}{The temperatures of the hot and cold flows on either side of the heat exchanger have a temperature difference known as an approach temperature.} A diagram showing a simplified counter-flow heat exchanger is illustrated in Figure \ref{counter-flow-hx}.

@@

Where $C_{avg}$ is the average capacitance rate across the hot or cold stream and $c_{in}$ and $c_{out}$ is the specific heat at the inlet and outlet respectively. \replaced[comment={Improving counter-flow hx explanation [1,1]}]{Specific heat is found using library correlations, with the average capacitance rate assumed to be constant during the analysis.}{Specific heat is found using library correlations.} Once both average capacitances are calculated for the hot and cold streams, one has a larger value, $\dot{C}_{max}$, and one has a smaller value,  $\dot{C}_{min}$. These maximum and minimum values are used to find the capacitance ratio, $CR$, in Equation \ref{cap_ratio}.

@@

\replaced[comment={Improving counter-flow hx explanation [1,1]}]{Assuming constant average capacitance rate is suitable for most engineering purposes, especially when there is uncertainty associated with other design parameters \cite{nellis_klein_2008}. To ensure that there is no internal pinch point, the temperatures of the hot and cold streams as a function of dimensionless length in the LTR and HTR are shown in Figure \ref{graph-t-vs-p-ltr} and Figure \ref{graph-t-vs-p-htr} respectively. Additionally, the specific heat as a function of dimensionless length of the hot and cold streams in the LTR and HTR are plotted in Figure \ref{graph-c-vs-p-ltr} and Figure \ref{graph-c-vs-p-htr} respectively.}{}

\begin{figure*}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{Definitions/c-vs-p-ltr.pdf}
        \caption[]%
        {{\small Specific heat as a function of location for LTR}}
        \label{graph-c-vs-p-ltr}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{Definitions/t-vs-p-ltr.pdf}
        \caption[]%
        {{\small Temperature as a function of location for LTR}}
        \label{graph-t-vs-p-ltr}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{Definitions/c-vs-p-htr.pdf}
        \caption[]%
        {{\small Specific heat as a function of location for HTR}}
        \label{graph-c-vs-p-htr}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{Definitions/t-vs-p-htr.pdf}
        \caption[]%
        {{\small Temperature as a function of location for HTR}}
        \label{graph-t-vs-p-htr}
    \end{subfigure}
    \caption{Specific heats and temperatures of hot and cold streams as a function of dimensionless location for the low temperature recuperator and high temperature recuperator}
    \label{graphs-t-c-vs-p} 
\end{figure*}

\replaced{The calculations used to discretize the counter-flow heat exchangers into sub-heat exchanger models is from JJ Dyreby's 2014 PhD thesis \cite{dyreby_2014}.}{}

@@

Where the left hand side, $\dot{m}$, $h_{inlet}$, and $\dot{Q}_{LFRHX}$, is the energy into the flow and the right hand side, $\dot{m}$ and $h_{outlet}$, is the energy brought out from the flow of sCO$_2$. The amount of energy transferred into the cycle, $\dot{Q}_{LFRHX}$, is set at $950$ MW, and outlet temperature of the sCO$_2$ from the LFR HX is set at a value of $595^{\circ}$C. The outlet temperature of the LFR is specified because of high temperature material limits on the LFR lead side. The low temperature side is allowed to vary over a range of values with some considerations. The lead flow velocity is limited by the erosion of the fuel, the slower the lead flow velocity reduces fuel erosion and therefore leads to a more desirable compact design. \replaced[comment={Edited sentence for clarity [2,3]}]{At constant lead velocity (and hence mass flow rate), reducing sCO$_2$ inlet temperature allows for a higher lead delta T in the LFR core and hence a higher thermal power output}{Restricting the lead flow velocity, and therefore lead mass flow rate, leads to a higher LFR power output when the inlet sCO$_2$ temperature is reduced.} LFR sCO$_2$ inlet temperature has a lower bound of $340^{\circ}$C before the lead begins to freeze, which is operationally unacceptable. When the inlet temperature of sCO$_2$ is increased the temperature difference across the LFR is decreased leading to an increase in power conversion cycle thermodynamic efficiency but a reduction in LFR power below $950$ MW. There is a compromise between high LFR efficiency and LFR power, therefore a temperature of $400^{\circ}$C for the sCO$_2$ inlet temperature is the optimal value provided by Westinghouse.

@@

When grid demand for electrical power increases, a series of operating modes are activated. During the highest demand times, cycle operation focuses on maximum electrical generation. This is achieved through the C2S being turned off for direct electrical production from the LFR and the hot TES is discharging heat through the CSP HX for electrical production in the sCO$_2$ Brayton cycle. \replaced[comment={Edited sentence to clarify that the CSP charges the hot TES during low grid demand periods [2,4]}]{As grid demand diminishes, CSP HX ramps down heat extraction until no power is being dispatched through the salt and the hot TES begins charging. During this process, the LFR gradually adds a larger fraction of heat input to the TES through C2S, supplementing the heat produced by the CSP which is also used to charge the TES. This process continues until no electrical production is occurring in the cycle and all heat is stored in TES for later use.}{As grid demand diminishes, CSP HX ramps down heat extraction until no power is being dispatched through the salt and the hot TES begins charging. During this process, the LFR gradually adds a larger fraction of heat input to the TES through C2S. This process continues until no electrical production is occurring in the cycle and all heat is stored in TES for later use.}
